Secure Composition of Robust and Optimising Compilers

### 2023-07-10 popl/main.tex
incorporate feedback and add example to sec 5.5

\begin{gather*}
  \begin{aligned}
  \mi{(Pre\text{-}Events)}~&\event[_{b}]\bnfdef \ev{Alloc}\ \loc\ \valueexpr \mid \ev{Dealloc}\ \loc \mid \ev{Get}\ \loc\ \valueexpr &\mid \ev{Set}\ \loc\ \valueexpr\ \valueexpr['] \mid \ev{Call}\ \comm\ g\ \valueexpr \mid \ev{Ret}\ \comm\ \valueexpr \mid \ev{Start} \mid \ev{End}\ \valueexpr \\
  \mi{(Events)}~&\event\bnfdef \emptyevent \mid \lightning \mid \tup{\event[_{b}]}{\sandboxtag} \\
  \end{aligned}
\end{gather*}

Possible events that are emitted during execution are similar to the ones of the specification trace model (\Cref{subsec:propdefs}).
The key difference is that, as standard in secure compilation work, the traces have a call and return event that signals context switches.\MP{why? to help with proofs?}
Hereby, a $\ev{Call\ \ctxtocomp\ foo\ \valueexpr}$ and $\ev{Return\ \ctxtocomp\ \valueexpr}$ signal that program execution transitions from context- to component-level.
Contrary, $\ev{Call\ \comptoctx\ foo\ \valueexpr}$ and $\ev{Return\ \ctxtocomp\ \valueexpr}$ signal that program execution transitions from component- to context-level.
The environmental semantics emits the tag $\nocomm$ for calls from context- to context-level or component- to component-level.
In the following, $\neg\ctx=\comp$ and $\neg\comp=\ctx$ and also $\operatorname{comm}(\ctx)=\ \ctxtocomp$ and $\operatorname{comm}(\comp)=\ \comptoctx$.
\MP{intuition? these are aux functions for the semantics}

@@

$\src{L_{\tmssafe}}$ does not extend the base syntax at term level.\MP{avoid negations. ltms programs use the same syntax presentd before without extensions}
What makes $\src{L_{\tmssafe}}$ special is that it has a type system inspired by $L^{3}$~\cite{morrisett2005L3} that exhibits the property that every well-typed $\src{L_{\tmssafe}}$ program satisfies \gls{tms}.\MPin{it's not really special ... Just say  "L3 has a type system ... ""}
The proof of this property relies on a projection $\msfilterLtms{\src{\event}}=\event$ that projects $\src{L_{\tmssafe}}$ events to specfication events $\event$. \MP{why}
Hereby, $\delta$ maps $\src{L_{\tmssafe}}$ locations $\src{\loc}$ to locations $\loc$ of the specification trace model.\MP{why, where is this needed}
The projection $\msfilterLtms{\src{\event}}$ for the most part does the expected\MP{what}.
But, it also filters any action that a contexts does, since these are irrelevant for component-level \gls{tms}.

@@

$\Ltrg$ extends the syntax presented earlier (\Cref{subsec:cs:defs}) with dynamic typechecks and a way to inspect poison tags in the metadata of pointers.
If an object\MP{?} a pointer bound to variable $\trg{x}$ points to is allocated ($\trg{\poisonless}$), the check $\trg{\lispoisoned{x}}$ yields $\trg{1}$.
Likewise, if it\MP{what} has been poisoned ($\trg{\poisoned}$), the check $\trg{\lispoisoned{x}}$ evaluates to $\trg{0}$.
Besides these checks, the language remains exactly the same\MP{to what}.

The projection $\msfilterL{\trg{\trace}}$ is entirely similar to $\msfilterLtms{\src{\trace}}$.

\subsection{$\Lms$: Another Memory-Unsafe Language}\label{subsec:lms}
This paper chooses to introduce\MP{uses / defines . use 1 word} $\Lms$, despite it being exactly equal to $\Ltrg$.
The main purpose of this decision is to enhance readability.\MPin{you can shorten these 2 sentences into 1. yes the section will be supershort, it's ok}
The projection $\msfilterLms{\irl{\trace}}$ is also exactly equal to $\msfilterL{\trg{\trace}}$.

\subsection{$\Lscct$: An Unsafe Language}\label{subsec:lscct}
\MPin{
	unsafe? 
	a possibly-CCT language?
}

$\Lscct$ extends the dynamic state\MPin{i really prefer operational state ... or runtime state. i think i'm not used to this terminology.} of $\trg{L}$ (\Cref{subsec:lms}) with a register for a data (operand) independent timing mode.
For the other languages seen earlier, this mode is intuitively always-on.
\MP{spell out the formal changes in an align* environmtn}
To modify the mode, $\Lscct$ extends expressions with the term $\obj{\lwrdoit{\expr}}$ that writes the value of $\obj{\expr}$ into the register and with a term $\obj{\lrddoit{x}{\expr}}$ that reads the value of the register into $\obj{x}$.
If the register is turned on, no secrets can appear on traces.
If the register is turned off, secrets may appear on traces as well as new events $\obj{Branch\ n}$ and $\obj{Binop\ n}$, which the semantics emits for conditionals and certain binary operations, e.g., division.
To prevent leakage of secrets but still enable reasoning about memory safety, pre-events are extended with $\obj{\widehat{{Get}}\ \loc\ \valueexpr}$ and $\obj{\widehat{Set}\ \loc\ \valueexpr}$, which indicate reads from and writes to memory without leaking secret information involved in the access.
\MPin{
	an example program and how it uses the flag and what kind of traces are produced is really needed here
}

The filters to the specification trace model are straightforward.

\section{Case Study: Composing Secure Compiler Passes and Optimisations \pages{7}}\label{sec:casestud:rtp}
\MP{use modifer [!h]. i HATE figs that float outside their section}
\begin{figure}[!h]
  \begin{center}
    \begin{tikzpicture}
      \node (S) {$\src{L_{\tmssafe}}$};
      \node[right=1.5 of S] (T) {$\trg{L}$};
      \node[right=1.5 of T] (M) {$\irl{L_{\mssafe}}$};
      \node[below right=0.5 and 1.0 of M] (D0) {$\irl{L_{\mssafe}}$};
      \node[above right=0.5 and 1.0 of M] (C0) {$\irl{L_{\mssafe}}$};
      \node[right=3.0 of M] (E) {$\irl{L_{\mssafe}}$};
      \node[right=1.5 of E] (O) {$\obj{L_{\scctsafe}}$};

>>>>

\begin{gather*}
  \begin{aligned}
  \mi{(Pre\text{-}Events)}~\event[_{b}]&\bnfdef \ev{Alloc}\ \loc\ \valueexpr \mid \ev{Dealloc}\ \loc \mid \ev{Get}\ \loc\ \valueexpr \mid \ev{Set}\ \loc\ \valueexpr\ \valueexpr['] \mid \ev{Call}\ \comm\ g\ \valueexpr \mid \ev{Ret}\ \comm\ \valueexpr \mid \ev{Start} \mid \ev{End}\ \valueexpr \\
  \mi{(Events)}~\event&\bnfdef \emptyevent \mid \lightning \mid \tup{\event[_{b}]}{\sandboxtag} \\
  \end{aligned}
\end{gather*}

Possible events that are emitted during execution are similar to the ones of the specification trace model (\Cref{subsec:propdefs}).
The key difference is that, as standard in secure compilation work~\cite{patrignani2021rsc,abate2019jour}, the traces have a call and return event that signals context switches, which are referred to as {\em interaction events}.
The reason for these interaction events is technical: They are a proof artifact for reconstructing a source context from a potentially malicious target context, where during that translation, the insertion of some wrapper code right before context switching may be necessary to make the proof succeed.
Hereby, a $\ev{Call\ \ctxtocomp\ foo\ \valueexpr}$ and $\ev{Return\ \ctxtocomp\ \valueexpr}$ signal that program execution transitions from context- to component-level.
Contrary, $\ev{Call\ \comptoctx\ foo\ \valueexpr}$ and $\ev{Return\ \ctxtocomp\ \valueexpr}$ signal that program execution transitions from component- to context-level.
The environmental semantics emits the tag $\nocomm$ for calls from context- to context-level or component- to component-level.
In the following, $\neg\ctx=\comp$ and $\neg\comp=\ctx$.

@@

$\src{L_{\tmssafe}}$ uses the same syntax as presented earlier (\Cref{sec:casestud:defs}) without extensions to the term level.
But, $\src{L_{\tmssafe}}$ is statically typed, where the type system is inspired by $L^{3}$~\cite{morrisett2005L3}.
The type system of $\src{L_{\tmssafe}}$ exhibits the property that every well-typed $\src{L_{\tmssafe}}$ program satisfies \gls{tms}.
The proof of this property relies on a projection $\msfilterLtms{\src{\event}}=\event$ that projects $\src{L_{\tmssafe}}$ events to specfication events $\event$, because the properties defined earlier (\Cref{subsec:propdefs}) are defined in the specification trace model.
Hereby, $\delta$ maps $\src{L_{\tmssafe}}$ locations $\src{\loc}$ to locations $\loc$ of the specification trace model whenever an event with a location is projected.
The projection $\msfilterLtms{\src{\event}}$ for the most part does the expected, e.g., $\msfilterLtms{\src{\ev{Alloc}\ \loc\ n;\comp;\securitytag}}=\ev{Alloc}\ \delta(\loc)\ n;\comp;\securitytag$.
But, it also filters any action that a context does as well as the interaction events, since these are irrelevant for component-level \gls{tms}.

@@

$\Ltrg$ extends the syntax presented earlier (\Cref{subsec:cs:defs}) with dynamic typechecks and a way to inspect poison tags in the metadata of pointers.
For allocated ($\trg{\poisonless}$) arrays bound to variable $\trg{x}$, the check $\trg{\lispoisoned{x}}$ yields $\trg{1}$.
If the array bound to $\trg{x}$ was allocated, i.e., has been poisoned ($\trg{\poisoned}$), the check $\trg{\lispoisoned{x}}$ evaluates to $\trg{0}$.
%
The projection $\msfilterL{\trg{\trace}}$ is entirely similar to $\msfilterLtms{\src{\trace}}$.

\subsection{$\Lms$: Another Memory-Unsafe Language}\label{subsec:lms}
To enhance readablity, this paper uses $\Lms$, despite it being exactly equal to $\Ltrg$.
The projection $\msfilterLms{\irl{\trace}}$ is also exactly equal to $\msfilterL{\trg{\trace}}$.

\subsection{$\Lscct$: A Language with a Data Independent Timing Mode}\label{subsec:lscct}

While the previous languages (\Cref{subsec:ltms,subsec:lsms,subsec:lms}) are not concerned with \gls{cct}, $\Lscct$ extends $\Lms$ (\Cref{subsec:lms}) with a register flag for a data (operand) independent timing mode\footnote{As present in both Arm~\cite[p.~543]{arm-refman} and Intel~\cite[p.~80]{intel-refman} processors.}.
For the other languages seen earlier, this mode is intuitively always-on.
\MP{spell out the formal changes in an align* environmtn} % this necessitates that we present Ω also in the shared lang defs
To modify the mode, $\Lscct$ extends expressions with the term $\obj{\lwrdoit{\expr}}$ that writes the value of $\obj{\expr}$ into the register and with a term $\obj{\lrddoit{x}{\expr}}$ that reads the value of the register into $\obj{x}$.
If the register is turned on, no secrets can appear on traces.
If the register is turned off, secrets may appear on traces as well as new events $\obj{Branch\ n}$ and $\obj{Binop\ n}$, which the semantics emits for conditionals and certain binary operations, e.g., division.
To prevent leakage of secrets but still enable reasoning about memory safety, pre-events are extended with $\obj{\widehat{{Get}}\ \loc\ \valueexpr}$ and $\obj{\widehat{Set}\ \loc\ \valueexpr}$, which indicate reads from and writes to memory without leaking secret information involved in the access, which is taken care of by the projection to specification events.
\Cref{ex:lscct} illustrates these changes.

The projection to the specification trace model is mostly straightforward.
But for events containing the pre-events $\obj{\widehat{Get}}$ and $\obj{\widehat{Set}}$, the projection always translates the security-tag $\obj{\sigma}$ to $\unlock$, regardless of its actual value.
The pre-events themselves still translate to just $\ev{Get}$ and $\ev{Set}$, respectively.
With this technical setup, the information wether a read or write happened on a secret value is not hidden by the semantics, e.g., by emitting $\obj{\emptyevent}$, but when projecting to specification events.
This allows flexibility: The trace can be checked to satisfy different properties, such as, in this case, \gls{tms}, \gls{sms}, \gls{scct}, and their combined versions.

\begin{example}[$\Lscct$ with and without data independent timing]\label{ex:lscct}
  Consider the following $\Lscct$ program and the execution trace $\obj{\trace}$ juxtaposed with it.

  $$
  \begin{array}{lcc}
  \obj{\lwrdoit{0};} & \mid & \obj{\emptyevent}\\
  \obj{\llet{x_{\lock}}{\lnew{5}{4}}{}} & \mid & \obj{\ev{Alloc}\ \loc_{x}\ 4;\sandboxtag;\unlock} \\
  \obj{\lset{x}{2}{\lifz{0_{\lock}}{1}{3}};} & \mid & \obj{\ev{Branch}\ 0;\sandboxtag;\lock\cdot\ev{Set}\ \loc_{x}\ 2\ 1;\sandboxtag;\lock} \\
  \obj{\lwrdoit{1};} & \mid & \obj{\emptyevent}\\
  \obj{\llet{y_{\lock}}{\lnew{5}{4}}{}} & \mid & \obj{\ev{Alloc}\ \loc_{y}\ 4;\sandboxtag;\unlock}\\
  \obj{\lset{y_{\lock}}{2}{\lifz{0_{\lock}}{1}{3}};} & \mid & \obj{\emptyevent\cdot\widehat{Set}\ \loc_{y}\ 2;\sandboxtag;\lock}\\
  \end{array}
  $$

  When the data independent timing mode is off, the execution yields events in similar fashion to before (\Cref{subsec:ltms,subsec:lsms,subsec:lms}).
  But, if it is turned on, then the branching event does not fire anymore.
  Moreover, writing to memory now yields the pre-event $\obj{\widehat{Set}}$, whose difference to $\obj{Set}$ becomes apparent when looking at the projection to the specification trace model.
  Projecting the above trace to the specification trace model yields the following trace (emitting empty events):
  $\ev{Alloc}\ \loc_{x}\ 4;\sandboxtag;\unlock\cdot\ev{Branch\ 0}\cdot\ev{Set}\ \loc_{x}\ 2\ 1;\sandboxtag;\lock\cdot\ev{Alloc}\ \loc_{y}\ 4;\sandboxtag;\unlock\cdot\ev{Set}\ \loc_{y}\ 2;\sandboxtag;\unlock$.
  Clearly, secrets are hidden ($\unlock$) after this projection for the section of the program where the data independent timing mode is turned on, while still containing the appropriate events to check for \gls{ms}.
\end{example}

\section{Case Study: Composing Secure Compiler Passes and Optimisations \pages{7}}\label{sec:casestud:rtp}
\begin{figure}[!h]
  \begin{center}
    \begin{tikzpicture}
      \node (S) {$\src{L_{\tmssafe}}$};
      \node[right=1.5 of S] (T) {$\trg{L}$};
      \node[right=1.5 of T] (M) {$\irl{L_{\mssafe}}$};
      \node[below right=0.5 and 1.0 of M] (D0) {$\irl{L_{\mssafe}}$};
      \node[above right=0.5 and 1.0 of M] (C0) {$\irl{L_{\mssafe}}$};
      \node[right=3.0 of M] (E) {$\irl{L_{\mssafe}}$};
      \node[right=1.5 of E] (O) {$\obj{L_{\scctsafe}}$};
